---
title: "NIMBLE's version of the BUGS language (or, how to convert code from BUGS/JAGS to NIMBLE)"
author: "Perry de Valpine"
date: "September 2019"
output:
  slidy_presentation: default
  beamer_presentation: default
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)
```

How NIMBLE is the same as BUGS and JAGS
=====

* Most BUGS and JAGS distributions and functions are supported in NIMBLE

    - For distributions, see [User Manual Section 5.2.4](https://r-nimble.org/html_manual/cha-writing-models.html#subsec:dists-and-functions)
    - For function, see  [User Manual Section 5.2.5](https://r-nimble.org/html_manual/cha-writing-models.html#subsec:BUGS-lang-fxns)
	
* Most BUGS and JAGS syntax is supported in NIMBLE

    - Truncation syntax is different when using `nimbleCode`. 
    - `T(dnorm(mu, tau), 0, )` in NIMBLE instead of `dnorm(mu, tau) (0,)`.
    - It can be the same as for BUGS/JAGS if reading code from a file with `readBUGSmodel`).

Converting from JAGS/BUGS) to NIMBLE
=====

See our [guide online](https://r-nimble.org/quick-guide-for-converting-from-jags-or-bugs-to-nimble).

How NIMBLE extends BUGS
=====

- Alternative distribution parameterizations (like R).
- Named parameters (like R).
- Vectorized math and linear algebra
- Definition-time if-then-else (multiple model variants from the same code)
- User-defined functions and distributions
- Distinction between `constants` and `data`.

How NIMBLE is limited compared to BUGS and/or JAGS:
=====

- NIMBLE is stricter about requiring square brackets and informative indices for non-scalar nodes.

Alternative distribution parameterizations and named parameters
=====

- BUGS/JAGS: Only `dnorm(mu, tau)` is supported, where `tau` is precision.
- NIMBLE: Alternative parameterizations and named parameters are supported:

    - `dnorm(mean = mu, sd = sigma)`
    - `dnorm(mean = mu, var = sigma_squared)`
    - `dnorm(mean = mu, tau = tau)` (tau = precision)
    
- Distributions with alternative parameterizations are listed in Table 5.2 of [User Manual Section 5.2.4](https://r-nimble.org/html_manual/cha-writing-models.html#subsec:dists-and-functions)

Vectorized math and linear algebra
=====
More basic math is supported.  A lot of it is R-like.

- `x[1:5] <- (A[1:5, 1:5] %*% b[1:5] + c[1:5])[,1]`
- Need to be careful about indexing.  This expression returns a matrix, and we convert it to a vector with `[,1]` at the end.


Extensibility: new functions and distributions
=====

## Say we have an occupancy model.

- We want to write `y[i, 1:J] ~ dOcc(occProb = psi[i], detectionProb = p[i, 1:J])`
- The idea is to marginalize (sum probabilities) over the unknown occupancy state.
- This means the MCMC will not need to sample occupancy states (pro) but `dOcc` will involve more computation than `dbern` (con).
- We will see how to do this later.


Distinction between data and constants
=====

## Constants are values needed to define model relationships

- Index ranges like `N`
- Constant indexing vectors
- Constants must be provided when creating a model with `nimbleModel`.

## Data represents a flag on the role a parameter (node) plays in the model

- e.g. Data nodes shouldn't be sampled in MCMC.
- Data values can be changed.
- Data can be provided when calling `nimbleModel` or later.

## Providing data and constants together.

- Data and constants can be provided together **as `constants`**.
- It would be slightly easier for BUGS/JAGS users to call this "data", but that would blur the concepts.  They are more correctly viewed as `constants` if provided together.
- NIMBLE will usually disambiguate data when it is provided as constants.

# More explicit need to provide dimensions

- Square brackets must always be provided to indicate number of dimensions

    - If `x` is 2-dimensional, use `x[,] %*% beta[]`, not `x %*% beta`
    
* Sometimes NIMBLE is not as smart as BUGS/JAGS at determining dimensions.  There are two solutions:

    * Give dimensions explicity: `x[1:n, 1:m]`, OR
    * Provide a `dimensions` argument to `nimbleModel`.  Example: `dimensions = list(x = c(n, m))`.

# Vectorized math and linear algebra

Instead of writing this in your model code:
```{r, vectorized-1}
nimbleOptions(verbose = FALSE)
m1 <- nimbleModel(
    nimbleCode({
        for(i in 1:5) {
            predicted[i] <- beta0 + beta1 * x[i]
        }
    }
    ))
```
you can write this:
```{r, vectorized-2}
m2 <- nimbleModel(
    nimbleCode({
        predicted[1:5] <- beta0 + beta1 * x[1:5]
    }
    ))
```

These are not equivalent:
```{r, vectorized-3}
## m1 has 5 scalar nodes
m1$getNodeNames()
## m2 has 1 vector node
m2$getNodeNames()
```

One is not necessarily better than the other.  It depends on the model
and the MCMC configuration.  (More on those topics later.)

Vectorized declarations do not work for distributions.

Discussion topic for breaks:

- Would vectorized distribution declarations be helpful?
- How should they be defined?


# Definition-time if-then-else

If you wish to define multiple alternative models in one set of code,
you can use if-then-else statements.  These will be evaluated based on
variables in the R environment when the model is defined.  For
example:

```{r if-then-else}
code <- nimbleCode({
    sigma ~ dunif(0, 10)
    beta0 ~ dnorm(0, sd = 1000)
    beta1 ~ dnorm(0, sd = 1000)
    if(INCLUDE_X2) {beta2 ~ dnorm(0, sd = 1000)} else{}
    for(i in 1:10) {
        if(INCLUDE_X2) {
            y[i] ~ dnorm(beta0 + beta1 * x1[i] + beta2 * x2[i], sd = sigma)
        } else {
            y[i] ~ dnorm(beta0 + beta1 * x1[i], sd = sigma)
        }
    }
})

INCLUDE_X2 <- FALSE
m1 <- nimbleModel(code)
INCLUDE_X2 <- TRUE
m2 <- nimbleModel(code)
m1$getNodeNames()
m2$getNodeNames()

```

m2 has `beta2` while m1 does not.  The long names are "lifted nodes"
-- more on those later.

Discussion topic for breaks:

- What would be a more useful system for writing model variants within
  one block of code?
