---
title: "Strategies for improving MCMC"
author: "Perry de Valpine"
date: "September 2019"
output:
  slidy_presentation: default
  beamer_presentation: default
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE)
library(nimble)
library(compareMCMCs)
library(coda)
```

1. Re-write the model.

    - Think like a graph.
    - Reduce computation.
    
2. Customize sampler choices:

    - Try blocking correlated parameters.
    - Try sampling standard deviations on a log scale.
    - Try slice samplers instead of Metropolis-Hastings.
    - Try multiple samplers for slow-mixing parameters.
    
3. Advanced: write new samplers that take advantage of particular model structures.

These slides include step `1`.

Example: house martin state-space model
=====

This is from BPA (Kéry and Schaub, see references slides).

- House martin abundance estimates from 1990-2009
- 6 additional years of posterior prediction are included
- The model is density-independent.
- There are no covariates.
- Process noise and observation error and both normal on a log scale.
- This is really simple but we'll pretend it isn't.

Code and data
=====
```{r}
martin_code <- nimbleCode({
  # Priors and constraints
  logN.est[1] ~ dnorm(5.6, 0.01)    # Prior for initial population size
  mean.r ~ dnorm(1, 0.001)          # Prior for mean growth rate
  sigma.proc ~ dunif(0, 1)          # Prior for sd of state process
  sigma2.proc <- pow(sigma.proc, 2)
  tau.proc <- pow(sigma.proc, -2)
  sigma.obs ~ dunif(0, 1)           # Prior for sd of observation process
  sigma2.obs <- pow(sigma.obs, 2)
  tau.obs <- pow(sigma.obs, -2)
  
  # State process
  for (t in 1:(T-1)){
    r[t] ~ dnorm(mean.r, tau.proc)
    logN.est[t+1] <- logN.est[t] + r[t]
  }
  
  # Likelihood
  # Observation process 
  for (t in 1:T) {
    y[t] ~ dnorm(logN.est[t], tau.obs)
  }
  
  # Population sizes on real scale
  for (t in 1:T) {
    N.est[t] <- exp(logN.est[t])
  }
})

# Code from BPA book:
pyears <- 6 # Number of future years with predictions
hm <- c(271, 261, 309, 318, 231, 216, 208, 226, 195, 226, 233, 209, 
        226, 192, 191, 225, 245, 205, 191, 174, rep(NA, pyears))
year <- 1990:(2009 + pyears)

# Bundle data
martin_data <- list(y = log(hm), T = length(year))
## NIMBLE will handle y as data, T as a constant

# Initial values
martin_inits <- function(){
  list(sigma.proc = runif(1, 0, 1), mean.r = rnorm(1),
       sigma.obs = runif(1, 0, 1), 
       logN.est = c(rnorm(1, 5.6, 0.1), 
                    rep(NA, (length(year)-1))))
}

martin_model <- nimbleModel(martin_code,
                            constants = martin_data, 
                            inits = martin_inits())
```

Two ways to write a state-space model
=====

Two equivalent ways to write state-space models:

1. Process-noises are random variables.  States are deterministic given process noises. 
2. States are random variables.

This model uses approach `1`.

Think like a graph 1.
=====

What are the nodes in this model?

```{r}
martin_model$getNodeNames()
```

Think like a graph 2.
=====

What calculations are required to sample (or "update") `r[24]`?

```{r}
martin_model$getDependencies("r[24]")
```

Think like a graph 3.
=====

What calculations are required to sample `r[20]`?
```{r}
martin_model$getDependencies("r[20]")
```

What about `r[10]`?
What calculations are required to sample `r[10]`?
```{r}
martin_model$getDependencies("r[10]")
```

Think like a graph 4.
=====
We see that writing a state-space model this way will be very computationally costly.  Sampling the process noise at each time requires re-calculation of the model for all subsequent times.

Rewrite the model
=====
Here is a more efficient way to write a state-space model.

```{r}
martin_code_alt <- nimbleCode({
  # Priors and constraints
  logN.est[1] ~ dnorm(5.6, 0.01)       # Prior for initial population size
  mean.r ~ dnorm(1, 0.001)             # Prior for mean growth rate
  sigma.proc ~ dunif(0, 1)          # Prior for sd of state process
  sigma2.proc <- pow(sigma.proc, 2)
  tau.proc <- pow(sigma.proc, -2)
  sigma.obs ~ dunif(0, 1)           # Prior for sd of observation process
  sigma2.obs <- pow(sigma.obs, 2)
  tau.obs <- pow(sigma.obs, -2)
  
  # Likelihood
  # State process
  for (t in 1:(T-1)) {
    logN.est[t+1] ~ dnorm(logN.est[t] + mean.r, tau.proc)
  }
  
  # Observation process
  for (t in 1:T) {
    y[t] ~ dnorm(logN.est[t], tau.obs)
  }
  
  # Population sizes on real scale
  for (t in 1:T) {
    N.est[t] <- exp(logN.est[t])
  }
})
martin_model_alt <- nimbleModel(martin_code_alt,
                                constants = martin_data,
                                inits = martin_inits())
```

Now we can see how this makes a lighter dependency structure:

```{r}
martin_model_alt$getDependencies('logN.est[24]')
martin_model_alt$getDependencies('logN.est[20]')
martin_model_alt$getDependencies('logN.est[10]')
```

Updating each latent state (now population size, not process noise) requires calculations only one time-step later.

Compare performance
=====
```{r}
martin_orig <- compareMCMCs(
  list(code = martin_code,
       constants = martin_data,
       inits = martin_inits()),
  MCMCs = c("nimble", "jags"),
  MCMCcontrol = list(niter = 100000,
                     burnin = 10000)
)
martin_orig <- renameMCMC(martin_orig, "nimble_orig", "nimble")
martin_orig <- renameMCMC(martin_orig, "jags_orig", "jags")

martin_alt <- compareMCMCs(
  list(code = martin_code_alt,
       constants = martin_data,
       inits = martin_inits()),
  MCMCs = c("nimble", "jags"),
  MCMCcontrol = list(niter = 100000,
                     burnin = 10000)
)

martin_alt <- renameMCMC(martin_alt, "nimble_alt", "nimble")
martin_alt <- renameMCMC(martin_alt, "jags_alt", "jags")
```

Look at results
=====
```{r}
make_MCMC_comparison_pages(c(martin_orig, martin_alt),
                           dir = "martin_results",
                           modelName = "martin-state-space-model")
```

Results are [here](martin_results/martin-state-space-model.html)

We see:

- JAGS is more efficient than nimble.  We'll address that later.
- The alternative version of the model is more efficient for both JAGS and nimble.
- The worst-mixing parameters are the standard deviations (sigmas).

We can see the raw material of the results like this:

```{r}
martin_alt$jags_alt$metrics
martin_alt$nimble_alt$metrics
```