---
title: "Getting set up for TWS-AFS nimble workshop"
author: "Perry de Valpine"
date: "September 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Thanks for signing up for the workshop on hierarchical modeling with `nimble` at the 2019 TWS-AFS meeting.

This note covers what you need to do to be set up for the workshop.

You should have a computer with:

- a relatively up-to-date version of R.
- the following packges:

    - nimble
    - mcmcplots
    - devtools (just in case)

- a working C++ compiler (see below).
- ability to install additional R packages if necessary.

During the day of the workshop, you may need to install an additional R package or two.  If so, I'll give you instructions on that.

# Installing and checking a C++ compiler

`nimble` generates and compiles C++ on your machine.  You don't need to know anything about C++, but you do need to get it installed.

The best resource on this is Chapter 4 of the NIMBLE User Manual: [https://r-nimble.org/html_manual/cha-installing-nimble.html](https://r-nimble.org/html_manual/cha-installing-nimble.html).  (That text is a link, but this pdf doesn't automatically show it that way.)

Some additional notes:

- On linux, often you will have working compiler tools automatically.
- On OS X, you need to be sure you have XCode command-line tools installed.  If you have newly installed XCode, you might need to open it once to give it a one-time authorization.
- On Windows, be sure the checkbox to modify the PATH is checked when you install Rtools.exe.

## Test code

To see if your computer is set up correctly and your C++ compiler works for `nimble`, please try the following:

```{r eval = FALSE}
library(nimble)
m <- nimbleModel(nimbleCode({x ~ dnorm(0, 1)}), inits = list(x = 0))
cm <- compileNimble(m)
MCMC <- buildMCMC(m)
CMCMC <- compileNimble(MCMC, project = m)
CMCMC$run(100)
```

You should see output similar to the following:
```{r eval = TRUE}
library(nimble)
m <- nimbleModel(nimbleCode({x ~ dnorm(0, 1)}), inits = list(x = 0))
cm <- compileNimble(m)
MCMC <- buildMCMC(m)
CMCMC <- compileNimble(MCMC, project = m)
CMCMC$run(100)
```
 
# Problems?

If you have any problems, please email me.

Usually it works smoothly.

If it doesn't, it can be difficult to diagnose problems specific to one machine, but I will try.

