---
title: "Introduction to NIMBLE"
author: "Perry de Valpine"
date: "September 2019"
output:
  slidy_presentation: default
  beamer_presentation: default
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)
```

# What Is NIMBLE?

- A framework for hierarchical statistical models and methods.
- A nearly drop-in alternative to WinBUGS, OpenBUGS and JAGS.
- A configurable system for MCMC.
- A library of other methods.
- A model-generic programming system to write new analysis methods.
- Numerical Inference for statistical Models using Bayesian and Likelihood Estimation.

# First example: An occupancy model

* AHM = [Applied Hierarchical Modeling in Ecology, Vol. I](https://www.mbr-pwrc.usgs.gov/pubanalysis/keryroylebook/) by Marc Kéry and J. Andrew Royle. 2015. Elsevier.
* Most AHM examples have been converted to NIMBLE: [https://github.com/nimble-dev/AHMnimble](https://github.com/nimble-dev/AHMnimble)
* Thanks to Marc Kéry, Andy Royle, and Mike Meredith for permission to post modified versions of their code on GitHub.
* Occupancy example from section 10.4:

    - Simulated data
    - `M` sites.
    - Each site is visited `J` times.
    - `y[i, j]` is detection (`1`) or non-detection(`0`) for visit `j` to site `i`.
    - Explanatory variables:

        - `vegHt` = Vegetation height: logistic effect on occupancy probability
        - `wind` = Wind speed: logistic effect on occupancy probability

# Occupancy example: Load the package
```{r}
library(nimble)
```

# Occupancy example: Write the model code **in R**
- Slightly modified from AHM.
- Derived quantities are removed.
```{r echo = TRUE}
Section10p4_code <- nimbleCode({
  # Priors
  mean.p ~ dunif(0, 1)         # Detection intercept on prob. scale
  alpha0 <- logit(mean.p)      # Detection intercept
  alpha1 ~ dunif(-20, 20)      # Detection slope on wind
  mean.psi ~ dunif(0, 1)       # Occupancy intercept on prob. scale
  beta0 <- logit(mean.psi)     # Occupancy intercept
  beta1 ~ dunif(-20, 20)       # Occupancy slope on vegHt
  
  # Likelihood
  for (i in 1:M) {
    # True state model for the partially observed true state
    z[i] ~ dbern(psi[i])      # True occupancy z at site i
    logit(psi[i]) <- beta0 + beta1 * vegHt[i]
    for (j in 1:J) {
      # Observation model for the actual observations
      y[i,j] ~ dbern(p.eff[i,j])    # Detection-nondetection at i and j
      p.eff[i,j] <- z[i] * p[i,j]   # 'straw man' for WinBUGS
      logit(p[i,j]) <- alpha0 + alpha1 * wind[i,j]
    }
  }
  # Derived quantities are removed.
}
)
```

# Occupancy example: Simulate data
(This code is modified from AHM.  It is here for completeness.)
```{r}
DO_PLOT <- TRUE
if(!exists("DO_PLOT"))
  DO_PLOT <- FALSE
print("do plot")
print(DO_PLOT)
# Choose sample sizes and prepare obs. data array y
set.seed(1)                   # So we all get same data set
M <- 100                      # Number of sites
J <- 3                        # Number of presence/absence measurements
y <- matrix(NA, nrow = M, ncol = J) # to contain the obs. data

# Create a covariate called vegHt
vegHt <- sort(runif(M, -1, 1)) # sort for graphical convenience

# Choose parameter values for occupancy model and compute occupancy
beta0 <- 0                    # Logit-scale intercept
beta1 <- 3                    # Logit-scale slope for vegHt
psi <- plogis(beta0 + beta1 * vegHt) # Occupancy probability
# plot(vegHt, psi, ylim = c(0,1), type = "l", lwd = 3) # Plot psi relationship

# Now visit each site and observe presence/absence perfectly
z <- rbinom(M, 1, psi)        # True presence/absence

# Look at data so far
table(z)

# Plot the true system state
if(DO_PLOT) {
  par(mfrow = c(1, 3), mar = c(5,5,2,2), cex.axis = 1.5, cex.lab = 1.5)
  plot(vegHt, z, xlab="Vegetation height", ylab="True presence/absence (z)", frame = F, cex = 1.5)
  plot(function(x) plogis(beta0 + beta1*x), -1, 1, add=T, lwd=3, col = "red")
}

# Create a covariate called wind
wind <- array(runif(M * J, -1, 1), dim = c(M, J))

# Choose parameter values for measurement error model and compute detectability
alpha0 <- -2                        # Logit-scale intercept
alpha1 <- -3                        # Logit-scale slope for wind
p <- plogis(alpha0 + alpha1 * wind) # Detection probability
# plot(p ~ wind, ylim = c(0,1))     # Look at relationship

# Take J = 3 presence/absence measurements at each site
for(j in 1:J) {
  y[,j] <- rbinom(M, z, p[,j])
}
sum(apply(y, 1, max))               # Number of sites with observed presences

# Plot observed data and true effect of wind on detection probability
if(DO_PLOT) {
  plot(wind, y, xlab="Wind", ylab="Observed det./nondetection data (y)", frame = F, cex = 1.5)
  plot(function(x) plogis(alpha0 + alpha1*x), -1, 1, add=T, lwd=3, col = "red")
}
# Look at the data: occupancy, true presence/absence (z), and measurements (y)
cbind(psi=round(psi,2), z=z, y1=y[,1], y2=y[,2], y3=y[,3])

# Create factors
time <- matrix(rep(as.character(1:J), M), ncol = J, byrow = TRUE)
hab <- c(rep("A", 33), rep("B", 33), rep("C", 34))  # Must have M = 100

# Bundle and summarize data set
str( occupancy_data <- list(y = y, 
                            vegHt = vegHt,
                            wind = wind,
                            M = nrow(y),
                            J = ncol(y),
                            XvegHt = seq(-1, 1, length.out=100),
                            Xwind = seq(-1, 1, length.out=100)) )

# Initial values: must give for same quantities as priors given !
zst <- apply(y, 1, max)        # Avoid data/model/inits conflict
occupancy_inits <- function(){
  list(z = zst, 
       mean.p = runif(1), 
       alpha1 = runif(1), 
       mean.psi = runif(1), 
       beta1 = runif(1))
}

```


Occupancy Example: Steps to use NIMBLE:
===============

1. Build the model.  It is an R object.
2. Build the MCMC.

    - 2a. Configure the MCMC.
    - 2b. Customize the MCMC.
    - 2c. Build the MCMC.

3. Compile the model and MCMC.
4. Run the MCMC.
5. Extract the samples.

- `nimbleMCMC` does all of this at once.
- Steps 2a-2c can be combined if no customization is needed.
- `runMCMC` does steps 4-5

    - manages burn-in, multiple chains, and extracting samples.
    
Occupancy Example: 1. Build the model
=====

```{r}
occModel <- nimbleModel(Section10p4_code,
                        constants = occupancy_data, #discuss "constants" vs. "data"
                        inits = occupancy_inits())
```

Occupancy Example: 2. Build the MCMC
=====
```{r}
occMCMC <- buildMCMC(occModel)
```

Occupancy Example: 3. Compile the model and MCMC.
=====

This can be done in one step or two.  We'll use two.

```{r}
CoccModel <- compileNimble(occModel)
CoccMCMC <- compileNimble(occMCMC, project = occModel)
```
Occupancy Example: 4. Run the MCMC
=====
```{r}
CoccMCMC$run(10000)
```


Occupancy Example: 5. Extract the samples
=====
```{r}
samples <- as.matrix(CoccMCMC$mvSamples)
```

Occupancy Example: Look at results:
=====

There are many packages for summarizing and plotting MCMC samples.  NIMBLE does not try to re-invent these wheels.

1. `mcmcplots`

```{r eval = FALSE}
library(mcmcplots)
mcmcplot(samples, dir = ".", filename = "occupancy_samples_mcmcplot")
```

```{r echo=FALSE}
library(mcmcplots)
mcmcplot(samples, dir = ".", filename = "orig_occupancy_samples_mcmcplot")
```

Results that comes with these slides are [here](orig_occupancy_samples_mcmcplot.html).

Results if you generated your own will be [here](occupancy_samples_mcmcplot.html).

2. `coda`

```{r eval = FALSE}
library(coda)
pdf("occupancy_samples_coda.pdf")
plot(as.mcmc(samples))
dev.off()
```

```{r echo = FALSE}
library(coda)
pdf("orig_occupancy_samples_coda.pdf")
plot(as.mcmc(samples))
dev.off()
```

Results that comes with these slides are [here](orig_occupancy_samples_coda.pdf).

Results if you generated your own will be [here](occupancy_samples_coda.pdf).

MCMC workflow in NIMBLE
=====

Here is a summary of the MCMC workflow in NIMBLE.

[nimbleWorkflow.pdf](nimbleWorkflow.pdf)

Occupancy Example: Doing it all at once with `nimbleMCMC`
=====

Start from:

- code
- constants + data
- inits

```{r}
samples2 <- nimbleMCMC(Section10p4_code,
                       constants = occupancy_data,
                       inits = occupancy_inits,
                       niter = 10000,
                       nburnin = 1000,
                       nchains = 2,
                       samplesAsCodaMCMC = TRUE)
summary(samples2) ## from coda
```

Occupancy Example: Managing the run with `runMCMC`
=====

Start from:

- Compiled MCMC

```{r}
samples3 <- runMCMC(CoccMCMC, 
                    niter = 10000,
                    nburnin = 1000,
                    nchains = 2,
                    samplesAsCodaMCMC = TRUE)
summary(samples3)
```

Other notes
=====

- `readBUGSmodel` will read BUGS/JAGS model code and variables from their standard file formats.

