---
title: "Customizing the occupancy model"
author: "Perry de Valpine"
date: "September 2019"
output:
  slidy_presentation: default
  beamer_presentation: default
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)
```

```{r echo=FALSE}
library(nimble)
```

Set up the example again:
=====
(This code is modified from AHM.  It is here for completeness.)
```{r}
DO_PLOT <- TRUE
if(!exists("DO_PLOT"))
  DO_PLOT <- FALSE
print("do plot")
print(DO_PLOT)
# Choose sample sizes and prepare obs. data array y
set.seed(1)                   # So we all get same data set
M <- 100                      # Number of sites
J <- 3                        # Number of presence/absence measurements
y <- matrix(NA, nrow = M, ncol = J) # to contain the obs. data

# Create a covariate called vegHt
vegHt <- sort(runif(M, -1, 1)) # sort for graphical convenience

# Choose parameter values for occupancy model and compute occupancy
beta0 <- 0                    # Logit-scale intercept
beta1 <- 3                    # Logit-scale slope for vegHt
psi <- plogis(beta0 + beta1 * vegHt) # Occupancy probability
# plot(vegHt, psi, ylim = c(0,1), type = "l", lwd = 3) # Plot psi relationship

# Now visit each site and observe presence/absence perfectly
z <- rbinom(M, 1, psi)        # True presence/absence

# Look at data so far
table(z)

# Plot the true system state
if(DO_PLOT) {
  par(mfrow = c(1, 3), mar = c(5,5,2,2), cex.axis = 1.5, cex.lab = 1.5)
  plot(vegHt, z, xlab="Vegetation height", ylab="True presence/absence (z)", frame = F, cex = 1.5)
  plot(function(x) plogis(beta0 + beta1*x), -1, 1, add=T, lwd=3, col = "red")
}

# Create a covariate called wind
wind <- array(runif(M * J, -1, 1), dim = c(M, J))

# Choose parameter values for measurement error model and compute detectability
alpha0 <- -2                        # Logit-scale intercept
alpha1 <- -3                        # Logit-scale slope for wind
p <- plogis(alpha0 + alpha1 * wind) # Detection probability
# plot(p ~ wind, ylim = c(0,1))     # Look at relationship

# Take J = 3 presence/absence measurements at each site
for(j in 1:J) {
  y[,j] <- rbinom(M, z, p[,j])
}
sum(apply(y, 1, max))               # Number of sites with observed presences

# Plot observed data and true effect of wind on detection probability
if(DO_PLOT) {
  plot(wind, y, xlab="Wind", ylab="Observed det./nondetection data (y)", frame = F, cex = 1.5)
  plot(function(x) plogis(alpha0 + alpha1*x), -1, 1, add=T, lwd=3, col = "red")
}
# Look at the data: occupancy, true presence/absence (z), and measurements (y)
cbind(psi=round(psi,2), z=z, y1=y[,1], y2=y[,2], y3=y[,3])

# Create factors
time <- matrix(rep(as.character(1:J), M), ncol = J, byrow = TRUE)
hab <- c(rep("A", 33), rep("B", 33), rep("C", 34))  # Must have M = 100

# Bundle and summarize data set
str( occupancy_data <- list(y = y, 
                            vegHt = vegHt,
                            wind = wind,
                            M = nrow(y),
                            J = ncol(y),
                            XvegHt = seq(-1, 1, length.out=100),
                            Xwind = seq(-1, 1, length.out=100)) )

# Initial values: must give for same quantities as priors given !
zst <- apply(y, 1, max)        # Avoid data/model/inits conflict
occupancy_inits <- function(){
  list(z = zst, 
       mean.p = runif(1), 
       alpha1 = runif(1), 
       mean.psi = runif(1), 
       beta1 = runif(1))
}
```

Original code
=====

```{r}
Section10p4_code <- nimbleCode({
  # Priors
  mean.p ~ dunif(0, 1)         # Detection intercept on prob. scale
  alpha0 <- logit(mean.p)      # Detection intercept
  alpha1 ~ dunif(-20, 20)      # Detection slope on wind
  mean.psi ~ dunif(0, 1)       # Occupancy intercept on prob. scale
  beta0 <- logit(mean.psi)     # Occupancy intercept
  beta1 ~ dunif(-20, 20)       # Occupancy slope on vegHt
  
  # Likelihood
  for (i in 1:M) {
    # True state model for the partially observed true state
    z[i] ~ dbern(psi[i])      # True occupancy z at site i
    logit(psi[i]) <- beta0 + beta1 * vegHt[i]
    for (j in 1:J) {
      # Observation model for the actual observations
      y[i,j] ~ dbern(p.eff[i,j])    # Detection-nondetection at i and j
      p.eff[i,j] <- z[i] * p[i,j]   # 'straw man' for WinBUGS
      logit(p[i,j]) <- alpha0 + alpha1 * wind[i,j]
    }
  }
  # Derived quantities are removed.
}
)
```

Use a new distribution to marginalize over `z[i]`
=====

```{r}
occCode2 <- nimbleCode({
  # Priors
  mean.p ~ dunif(0, 1)         # Detection intercept on prob. scale
  alpha0 <- logit(mean.p)      # Detection intercept
  alpha1 ~ dunif(-20, 20)      # Detection slope on wind
  mean.psi ~ dunif(0, 1)       # Occupancy intercept on prob. scale
  beta0 <- logit(mean.psi)     # Occupancy intercept
  beta1 ~ dunif(-20, 20)       # Occupancy slope on vegHt
  
  # Likelihood
  for (i in 1:M) {
    # True state model for the partially observed true state
    logit(psi[i]) <- beta0 + beta1 * vegHt[i]
    y[i, 1:J] ~ dOcc_v(probOcc = psi[i], probDetect = p[i, 1:J], len = J)
    for (j in 1:J) {
      logit(p[i,j]) <- alpha0 + alpha1 * wind[i,j]
    }
  }
}
)
```

How to write a new distribution: dOcc
=====

Prob(Detection history) = Prob(Detection history | Occupied) Prob(Occupied) + Prob(Detection history | Unoccupied) Prob(Unoccupied)

Hybrid math / code expression:
\[
P(y[i,1:J] | \psi, p[i, 1:J]) = \psi \prod_{j=1}^J \mbox{dbern}(y[i,j] | p[i,j]) + (1-\psi) I(\mbox{all } y[i, j] \mbox{ = 0})
\]

```{r}
dOcc_v <- nimbleFunction(
  run = function(x = double(1),
                 probOcc = double(0),
                 probDetect = double(1),
                 len = integer(0, default = 0),
                 log = logical(0, default = 0)) {
    if (len != 0) 
      if (len != length(x))
        stop("Argument 'len' must match length of data, or be 0.")
    if (length(x) != length(probDetect))
      stop("Length of data does not match length of detection vector.")
    returnType(double(0))
    logProb_x_given_occupied <- sum(dbinom(x,
                                           prob = probDetect, 
                                           size = 1,
                                           log = TRUE))
    prob_x_given_unoccupied <- sum(x) == 0
    prob_x <- exp(logProb_x_given_occupied) * probOcc + 
      prob_x_given_unoccupied * (1 - probOcc)
    if (log)
      return(log(prob_x))
    return(prob_x)
  }
)
```

Run MCMC with the new version:
=====

```{r}
occModel2 <- nimbleModel(occCode2,
                         constants = occupancy_data,
                         inits = occupancy_inits())
occMCMC2 <- buildMCMC(occModel2)
CoccModel2 <- compileNimble(occModel2)
CoccMCMC2 <- compileNimble(occMCMC2, project = occModel2)
samples <- runMCMC(CoccMCMC2, niter = 10000, nburnin = 1000)
```

Look at the results:
=====
```{r eval = FALSE}
library(mcmcplots)
mcmcplot(samples, dir = ".", filename = "occModel2_samples_mcmcplot")
```

```{r echo=FALSE}
library(mcmcplots)
mcmcplot(samples, dir = ".", filename = "orig_occModel2_samples_mcmcplot")
```

Results that comes with these slides are [here](orig_occModel2_samples_mcmcplot.html).

Results if you generated your own will be [here](occModel2_samples_mcmcplot.html).

Vectorize some calculations
=====

Vectorizing some calculations:

- Can make code more compact.
- Can make model and MCMC building and compiling faster.
- Can improve MCMC efficiency, but sometimes not by much.
- Can hurt MCMC efficiency if done in the wrong places.
- We'll learn to think about models more later.

```{r}
occCode3 <- nimbleCode({
  # Priors
  mean.p ~ dunif(0, 1)         # Detection intercept on prob. scale
  alpha0 <- logit(mean.p)      # Detection intercept
  alpha1 ~ dunif(-20, 20)      # Detection slope on wind
  mean.psi ~ dunif(0, 1)       # Occupancy intercept on prob. scale
  beta0 <- logit(mean.psi)     # Occupancy intercept
  beta1 ~ dunif(-20, 20)       # Occupancy slope on vegHt
  
  # Likelihood
  logit(p[1:M,1:J]) <- alpha0 + alpha1 * wind[1:M,1:J]
  for (i in 1:M) {
    # True state model for the partially observed true state
    z[i] ~ dbern(psi[i])      # True occupancy z at site i
    logit(psi[i]) <- beta0 + beta1 * vegHt[i]
    p.eff[i, 1:J] <- z[i] * p[i, 1:J]   # 'straw man' for WinBUGS
    for (j in 1:J) {
      # Observation model for the actual observations
      y[i,j] ~ dbern(p.eff[i,j])    # Detection-nondetection at i and j
    }
  }
  # Derived quantities are removed.
}
)
```

Run MCMC with the vectorized version:
=====

```{r}
occModel3 <- nimbleModel(occCode3,
                         constants = occupancy_data,
                         inits = occupancy_inits())
occMCMC3 <- buildMCMC(occModel3)
CoccModel3 <- compileNimble(occModel3)
CoccMCMC3 <- compileNimble(occMCMC3, project = occModel3)
samples <- runMCMC(CoccMCMC3, niter = 10000, nburnin = 1000)
```

Look at the results:
=====
```{r eval = FALSE}
library(mcmcplots)
mcmcplot(samples, dir = ".", filename = "occModel3_samples_mcmcplot")
```

```{r echo=FALSE}
library(mcmcplots)
mcmcplot(samples, dir = ".", filename = "orig_occModel3_samples_mcmcplot")
```

Results that comes with these slides are [here](orig_occModel3_samples_mcmcplot.html).

Results if you generated your own will be [here](occModel3_samples_mcmcplot.html).

