---
title: 'Example: Multi-state capture-recapture / HMM (Orchids)'
author: "Perry de Valpine"
date: "9/21/2019"
date: "September 2019"
output:
  slidy_presentation: default
  beamer_presentation: default
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)
library(nimble)
library(compareMCMCs)
```

Orchids (ladyslipper) example
=====

- From Bayesian Population Analysis Using WinBUGS (K&#233;ry and Schaub), aka BPA.
- Orchids modeled in states "vegetative", "flowering", "dormant", "dead"
- Observations: "seen vegetative", "seen flowering", "not seen".
- k = 11 annual observation periods 
- n = 250 unique plants
- n = 236 plants seen before the last observation time

This is a simple multi-state model because the state when not seen is unknown.

Obviously multi-state models can be more complicated than this.


Setup data
=====
```{r}
CH <- as.matrix(read.table("orchids.txt", sep=" ", header = FALSE))
head(CH) ## Glimpse of data
n_occasions <- dim(CH)[2]
## Compute vector with occasion of first capture
f <- numeric()
for (i in 1:dim(CH)[1])
  f[i] <- min(which(CH[i,]!=0))

## Modification for NIMBLE:
CH <- CH[which(f!=11), ]  ## remove all individuals not seen until the last occasion: They contribute no information and create problems for the dDHMM version.
## Reset f from the reduced CH
f <- numeric()
for (i in 1:dim(CH)[1])
  f[i] <- min(which(CH[i,]!=0))

## Recode CH matrix: note, a 0 is not allowed by WinBUGS!
## 1 = seen vegetative, 2 = seen flowering, 3 = not seen 
rCH <- CH  # Recoded CH 
rCH[rCH==0] <- 3 

## Function to create known latent states z 
known_state_ms <- function(ms, notseen){
    ## notseen: label for not seen 
    state <- ms 
    state[state==notseen] <- NA 
    for (i in 1:dim(ms)[1]){
        m <- min(which(!is.na(state[i,]))) 
        state[i,m] <- NA 
    }
    return(state) 
}
## Bundle data 
orchids_data <- list(y = rCH,
                  f = f,
                  n_occasions = dim(rCH)[2],
                  nind = dim(rCH)[1],
                  z = known_state_ms(rCH, 3)) 

ms_init_z <- function(ch, f){
   for (i in 1:dim(ch)[1]){ch[i,1:f[i]] <- NA}
   states <- max(ch, na.rm = TRUE)
   known.states <- 1:(states-1)
   v <- which(ch==states)
   ch[-v] <- NA
   ch[v] <- sample(known.states, length(v), replace = TRUE)
   return(ch)
   }

## Initial values 
orchids_inits <- function(){
    list(s = runif((dim(rCH)[2]-1), 0, 1),
         z = ms_init_z(rCH, f)
         )}

```


Model code
=====

- `ps[1, i, t, 2]` = probability of transition from state 1 to 2 for individual `i` from time `t` to time `t+1`.
- `po[1, i, t, 2]` = probability of being observed in state 2 when in state 1 for individual `i` at time `t`.
- `s[t]` = survival probability at time `t`.
- `psiV[j]` = probability of transition from state 1 (Vegetative) to state j.
- `psiF[j]` = probability of transition from state 2 (Flowering) to state j.
- `psiD[j]` = probability of transition from state 3 (Dormant) to state j.

```{r}
orchids_code <- nimbleCode({
  ## -------------------------------------------------
  ## Parameters:
  ## s: survival probability 
  ## psiV: transitions from vegetative 
  ## psiF: transitions from flowering 
  ## psiD: transitions from dormant 
  ## -------------------------------------------------
  ## States (S):
  ## 1 vegetative 
  ## 2 flowering 
  ## 3 dormant 
  ## 4 dead 
  ## Observations (O):  
  ## 1 seen vegetative 
  ## 2 seen flowering 
  ## 3 not seen 
  ## -------------------------------------------------
  ## Priors and constraints 
  ## Survival: uniform 
  for (t in 1:(n_occasions-1)){  
    s[t] ~ dunif(0, 1) 
  }
  ## Transitions: gamma priors 
  for (i in 1:3){
    a[i] ~ dgamma(1, 1) 
    psiD[i] <- a[i]/sum(a[1:3]) 
    b[i] ~ dgamma(1, 1) 
    psiV[i] <- b[i]/sum(b[1:3]) 
    c[i] ~ dgamma(1, 1) 
    psiF[i] <- c[i]/sum(c[1:3]) 
  }
  ## Define state-transition and observation matrices 	
  for (i in 1:nind){
    ## Define probabilities of state S(t+1) given S(t) 
    for (t in 1:(n_occasions-1)){
      ps[1,i,t,1] <- s[t] * psiV[1]
      ps[1,i,t,2] <- s[t] * psiV[2]
      ps[1,i,t,3] <- s[t] * psiV[3]
      ps[1,i,t,4] <- 1-s[t]
      ps[2,i,t,1] <- s[t] * psiF[1]
      ps[2,i,t,2] <- s[t] * psiF[2]
      ps[2,i,t,3] <- s[t] * psiF[3]
      ps[2,i,t,4] <- 1-s[t]
      ps[3,i,t,1] <- s[t] * psiD[1]
      ps[3,i,t,2] <- s[t] * psiD[2]
      ps[3,i,t,3] <- s[t] * psiD[3]
      ps[3,i,t,4] <- 1-s[t]
      ps[4,i,t,1] <- 0 
      ps[4,i,t,2] <- 0 
      ps[4,i,t,3] <- 0 
      ps[4,i,t,4] <- 1 
      ## Define probabilities of O(t) given S(t) 
      po[1,i,t,1] <- 1 
      po[1,i,t,2] <- 0 
      po[1,i,t,3] <- 0 
      po[2,i,t,1] <- 0 
      po[2,i,t,2] <- 1 
      po[2,i,t,3] <- 0 
      po[3,i,t,1] <- 0 
      po[3,i,t,2] <- 0 
      po[3,i,t,3] <- 1 
      po[4,i,t,1] <- 0 
      po[4,i,t,2] <- 0 
      po[4,i,t,3] <- 1 
    } #t 
  } #i 
  ## Likelihood
  for (i in 1:nind){
    ## Define latent state at first capture 
    z[i,f[i]] <- y[i,f[i]]
    for (t in (f[i]+1):n_occasions){
      ## State process: draw S(t) given S(t-1)
      z[i,t] ~ dcat(ps[z[i,t-1], i, t-1, 1:4]) 
      ## Observation process: draw O(t) given S(t) 
      y[i,t] ~ dcat(po[z[i,t], i, t-1, 1:3]) 
    } #t 
  } #i
})

orchids_info <- list(code=orchids_code, constants=orchids_data, inits=orchids_inits())
```

Run basic version
=====
```{r}
orchids_result <- compareMCMCs(
  orchids_info,
  MCMCs = c("nimble", "jags"),
  monitors = c("s", "psiV", "psiF", "psiD"),
  MCMCcontrol = list(niter = 20000,
                     burnin = 2000)
)
orchids_result <- renameMCMC(orchids_result, "nimble_orig", "nimble")
orchids_result <- renameMCMC(orchids_result, "jags_orig", "jags")
```

Version without unnecessary (computationally wasteful) indexing over individuals and/or time
=====
```{r}
#nimbleOptions(disallow_multivariate_argument_expressions = FALSE)
orchids_code2 <- nimbleCode({
  ## -------------------------------------------------
  ## Parameters:
  ## s: survival probability 
  ## psiV: transitions from vegetative 
  ## psiF: transitions from flowering 
  ## psiD: transitions from dormant 
  ## -------------------------------------------------
  ## States (S):
  ## 1 vegetative 
  ## 2 flowering 
  ## 3 dormant 
  ## 4 dead 
  ## Observations (O):  
  ## 1 seen vegetative 
  ## 2 seen flowering 
  ## 3 not seen 
  ## -------------------------------------------------
  ## Priors and constraints 
  ## Survival: uniform 
  for (t in 1:(n_occasions-1)){  
    s[t] ~ dunif(0, 1) 
  }
  ## Transitions: gamma priors 
  for (i in 1:3){
    a[i] ~ dgamma(1, 1) 
    psiD[i] <- a[i]/sum(a[1:3]) 
    b[i] ~ dgamma(1, 1) 
    psiV[i] <- b[i]/sum(b[1:3]) 
    c[i] ~ dgamma(1, 1) 
    psiF[i] <- c[i]/sum(c[1:3]) 
  }
  ## Define state-transition and observation matrices 	
  ## Define probabilities of state S(t+1) given S(t) 
  for (t in 1:(n_occasions-1)){
    ps[1,t,1] <- s[t] * psiV[1]
    ps[1,t,2] <- s[t] * psiV[2]
    ps[1,t,3] <- s[t] * psiV[3]
    ps[1,t,4] <- 1-s[t]
    ps[2,t,1] <- s[t] * psiF[1]
    ps[2,t,2] <- s[t] * psiF[2]
    ps[2,t,3] <- s[t] * psiF[3]
    ps[2,t,4] <- 1-s[t]
    ps[3,t,1] <- s[t] * psiD[1]
    ps[3,t,2] <- s[t] * psiD[2]
    ps[3,t,3] <- s[t] * psiD[3]
    ps[3,t,4] <- 1-s[t]
    ps[4,t,1] <- 0 
    ps[4,t,2] <- 0 
    ps[4,t,3] <- 0 
    ps[4,t,4] <- 1 
  }
  ## Define probabilities of O(t) given S(t) 
  po[1,1] <- 1 
  po[1,2] <- 0 
  po[1,3] <- 0 
  po[2,1] <- 0 
  po[2,2] <- 1 
  po[2,3] <- 0 
  po[3,1] <- 0 
  po[3,2] <- 0 
  po[3,3] <- 1 
  po[4,1] <- 0 
  po[4,2] <- 0 
  po[4,3] <- 1 
  ## Likelihood
  for (i in 1:nind){
    ## Define latent state at first capture 
    z[i,f[i]] <- y[i,f[i]]
    for (t in (f[i]+1):n_occasions){
      ## State process: draw S(t) given S(t-1)
      z[i,t] ~ dcat(ps[z[i,t-1], t-1, 1:4]) 
      ## Observation process: draw O(t) given S(t) 
      y[i,t] ~ dcat(po[z[i,t], 1:3]) 
    } #t 
  } #i
})

orchids_info2 <- list(code=orchids_code2, constants=orchids_data, inits=orchids_inits())
```

Run version without unnecessary indexing
=====
```{r}
orchids_result2 <- compareMCMCs(
  orchids_info2,
  MCMCs = c("nimble", "jags"),
  monitors = c("s", "psiV", "psiF", "psiD"),
  MCMCcontrol = list(niter = 20000,
                     burnin = 2000)
)
orchids_result2 <- renameMCMC(orchids_result2, "nimble_orig2", "nimble")
orchids_result2 <- enameMCMC(orchids_result2, "jags_orig2", "jags")
```

# DHMM version

```{r}
orchids_code_DHMM <- quote({
  ## -------------------------------------------------
  ## Parameters:
  ## s: survival probability 
  ## psiV: transitions from vegetative 
  ## psiF: transitions from flowering 
  ## psiD: transitions from dormant 
  ## -------------------------------------------------
  ## States (S):
  ## 1 vegetative 
  ## 2 flowering 
  ## 3 dormant 
  ## 4 dead 
  ## Observations (O):  
  ## 1 seen vegetative 
  ## 2 seen flowering 
  ## 3 not seen 
  ## -------------------------------------------------
  ## Priors and constraints 
  ## Survival: uniform 
  for (t in 1:(k-1)){  
    s[t] ~ dunif(0, 1)
  }
  ## Transitions: gamma priors 
  for (i in 1:3){
    a[i] ~ dgamma(1, 1) 
    psiD[i] <- a[i]/sum(a[1:3]) 
    b[i] ~ dgamma(1, 1) 
    psiV[i] <- b[i]/sum(b[1:3]) 
    c[i] ~ dgamma(1, 1) 
    psiF[i] <- c[i]/sum(c[1:3]) 
  }
  ## Define state-transition and observation matrices 	
  for (t in 1:(k-1)) {
    T[1,1,t] <- s[t] * psiV[1]
    T[2,1,t] <- s[t] * psiV[2]
    T[3,1,t] <- s[t] * psiV[3]
    T[4,1,t] <- 1-s[t]
    T[1,2,t] <- s[t] * psiF[1]
    T[2,2,t] <- s[t] * psiF[2]
    T[3,2,t] <- s[t] * psiF[3]
    T[4,2,t] <- 1-s[t]
    T[1,3,t] <- s[t] * psiD[1]
    T[2,3,t] <- s[t] * psiD[2]
    T[3,3,t] <- s[t] * psiD[3]
    T[4,3,t] <- 1-s[t]
    T[1,4,t] <- 0
    T[2,4,t] <- 0
    T[3,4,t] <- 0
    T[4,4,t] <- 1
  }
  O[1,1] <- 1 
  O[1,2] <- 0 
  O[1,3] <- 0 
  O[2,1] <- 0 
  O[2,2] <- 1 
  O[2,3] <- 0 
  O[3,1] <- 0 
  O[3,2] <- 0 
  O[3,3] <- 1 
  O[4,1] <- 0 
  O[4,2] <- 0 
  O[4,3] <- 1 
  for(i in 1:nind) {
    for(j in 1:4)
      init[i, j] <- y2[i, f[i]]==j ## y2 is the same as y, to avoid a cycle in graph (i.e. y[i, f[i]] depends on init[i, 1:3] which depends on y[i, f[i]])
  }
  for (i in 1:nind) {
    y[i, f[i]:k] ~ dDHMM(init = init[i, 1:4],
                         probObs = O[1:4,1:3],
                         probTrans = T[1:4,1:4,f[i]:(k-1)],
                         len = k-f[i]+1)
  }
})
orchids_constants_DHMM <- list(f=f,
                               k=dim(rCH)[2],
                               nind=dim(rCH)[1])
orchids_data_DHMM <- list(y = rCH, y2 = rCH)
orchids_inits_DHMM <- function() {
  list(s = runif((dim(rCH)[2]-1), 0, 1),
       a = rep(1,3),
       b = rep(1,3),
       c = rep(1,3))
}

orchids_info_DHMM <- list(code=orchids_code_DHMM, 
                          constants=orchids_constants_DHMM,
                          data = orchids_data_DHMM,
                          inits=orchids_inits_DHMM())

```


dDHMM from nimbleEcology
=====
```{r}
dDHMM <- nimbleFunction(
  run = function(x = double(1),    ## Observed capture (state) history
                 init = double(1),
                 probObs = double(2),
                 probTrans = double(3),
                 len = double(),## length of x (needed as a separate param for rDHMM)
                 log = integer(0, default = 0)) {
    if (length(init) != dim(probObs)[1]) stop("Length of init does not match nrow of probObs in dDHMM.")
    if (length(init) != dim(probTrans)[1]) stop("Length of init does not match dim(probTrans)[1] in dDHMM.")
    if (length(init) != dim(probTrans)[2]) stop("Length of init does not match dim(probTrans)[2] in dDHMM.")
    if (length(x) != len) stop("Length of x does not match len in dDHMM.")
    if (len - 1 != dim(probTrans)[3]) stop("len - 1 does not match dim(probTrans)[3] in dDHMM.")

    pi <- init # State probabilities at time t=1
    logL <- 0
    nObsClasses <- dim(probObs)[2]
    lengthX <- length(x)
    for (t in 1:lengthX) {
      if (x[t] > nObsClasses) stop("Invalid value of x[t] in dDHMM.")
      Zpi <- probObs[, x[t]] * pi # Vector of P(state) * P(observation class x[t] | state)
      sumZpi <- sum(Zpi)    # Total P(observed as class x[t])
      logL <- logL + log(sumZpi)  # Accumulate log probabilities through time
      if (t != lengthX) pi <- (probTrans[,,t] %*% asCol(Zpi) / sumZpi)[ ,1] # State probabilities at t+1
    }

    returnType(double())
    if (log) return(logL)
    return(exp(logL))
  }
)

rDHMM <- nimbleFunction(
  run = function(n = integer(),    ## Observed capture (state) history
                 init = double(1),
                 probObs = double(2),
                 probTrans = double(3),
                 len = double()) {
    nStates <- length(init)
    if (nStates != dim(probObs)[1]) stop("Length of init does not match nrow of probObs in dDHMM.")
    if (nStates != dim(probTrans)[1]) stop("Length of init does not match dim(probTrans)[1] in dDHMM.")
    if (nStates != dim(probTrans)[2]) stop("Length of init does not match dim(probTrans)[2] in dDHMM.")
    if (len - 1 != dim(probTrans)[3]) stop("len - 1 does not match dim(probTrans)[3] in dDHMM.")
    
    returnType(double(1))
    ans <- numeric(len)
    
    trueState <- rcat(1, init)
    for (i in 1:len) {
      # Detect based on the true state
      ans[i] <- rcat(1, probObs[trueState,])
      # Transition to a new true state
      if (i != len) {
        trueState <- rcat(1, probTrans[trueState, , i])
    }
  }
  return(ans)
})

registerDistributions(list(
  dDHMM = list(
    BUGSdist = "dDHMM(init, probObs, probTrans, len)",
    Rdist = "dDHMM(init, probObs, probTrans, len)",
    discrete = TRUE,
    types = c('value = double(1)',
              'init = double(1)',
              'probObs = double(2)',
              'probTrans = double(3)',
              'len = double()'),
    mixedSizes = TRUE,
    pqAvail = FALSE))
)
```

Run DHMM version
=====
```{r}
orchids_result_DHMM <- compareMCMCs(
  orchids_info_DHMM,
  MCMCs = c("nimble"),
  monitors = c("s", "psiV", "psiF", "psiD"),
  MCMCcontrol = list(niter = 20000,
                     burnin = 2000)
)
orchids_result_DHMM <- renameMCMC(orchids_result_DHMM, "nimble_DHMM", "nimble")
```

Make results figure
=====
```{r}
make_MCMC_comparison_pages(c(orchids_result, orchids_result2, orchids_result_DHMM),
                           modelName = "orchids")
```
